Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales

### 2020-03-24 reports/article.Rmd
version after international revision and improved figure resolution.

---
title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
author: Anne Garland`*`^1,2,3^, Erik Bulow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5^, Mark Wilkinson^6^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
date: '`r Sys.Date()`'
bibliography: library.bib
csl: the-lancet.csl
output: 
  bookdown::word_document2:
    reference_docx: article.dotx
    toc: false
    df_print: kable
---

@@

**OBJECTIVE:** Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality. However, no prediction tools based on European populations are available, diagnosis code-based instruments such as the Charlson co-morbidity index are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) class is imprecise. We thus developed a simple model to predict 90-day mortality after elective THA.

@@

**RESLUTS:** The unadjusted cumulative 90-day mortality was `r surv90d` in Sweden and `r njr_surv` in England and Wales. We propose a simple main effects model combining `r coefs_print_reduced`. This model had good discriminatory ability both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`, rendering it superioir to traditional models based on the ASA class `r get_auc("ASA")`, the Charlson co-morbitiy index `r get_auc("Charlson")`, and the Elixhauser com-morbidity index `r get_auc("Elixhauser")`. Our model was well calibrated for predicted probabilities up to 5 %. A web calculator to aid clinical usage was published at `r shiny_url`.

@@

Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery.[@Bozic2013a; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] Due to limited, often single-center based samples, a lack of reported calibration abilities, the absence of external validation and the use of extrapolation there is still no widely accepted model for the prediction of early mortality.[@Manning2016] Importantly, to our knowledge, no reported model has been developed or validated on a European population.

@@

We used data from the Swedish Hip Arthroplasty Register (SHAR) for model derivation and internal validation of discrimination and calibration. The best model was then validated externally on patients from England and Wales recorded in the National Joint Registry for England, Wales, Northern Ireland, the Isle of Man and the States of Guernsey (NJR). We focused on patients with cemented THA due to osteoarthritis. Cementation is the most common fixation technique used in Sweden, thus used as inclusion criteria to increase heterogeneity. Only the last operated hip was accounted for in patients with bilateral THA, since those are shown to better resemble unilateral THAs.[@Bulow2019a] (Figure \@ref(fig:flowchart)).

@@

The patient cohort was identified from the SHAR. The inclusion period started 2008 since this was the first year with the American Society for Anesthesiologists (ASA) class and Body Mass Index (BMI) systematically recorded. The last year was 2015, since we had access to co-morbidity from the national patient register up to this point. Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009] We used data from a variety of sources, as previously described by Cnudde et al.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, with a completeness of 96-98%.[@shpr2018] Data on educational level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) was recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status (married/un-married/divorced/widow[er]), were also collected from LISA. The Swedish National Patient Register (NPR) was used to assess co-morbidity during the year preceding index surgery. This register contains all relevant diagnoses coded by ICD-10, as well as dates of admission and discharge for in- and outpatient episodes in all private and public hospitals. Completeness for NPR is above 99 % and 85-95 % of all diagnostic codes are valid.[@Ludvigsson2011] Death dates were linked from the national population register.

@@

We used the Kaplan-Meier estimator to assess unadjusted mortality after cemented THA as background knowledge. Further analysis was based on logistic regression since no censoring occurred within the 90 day study period. We used a modelling procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020] Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1. Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process. 1,000 bootstrap samples were drawn from the observed data set.[@Austin2004] We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model. We then only kept $\lambda$:s minimizing the mean cross-validated deviances in each sample. Those $\lambda$:s were used to estimate model coefficients for each potential predictor. The magnitude (absolute values) of those estimates were used as a measure of variable importance. Piece-wise linear regression was used to detect a break point where a significant drop in variable importance was observed. Potential predictors with variable importance above this break point were considered important and kept as model candidates. The whole process was repeated 100 times. Covariates that were selected at least once out of the 100 times were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables (main model). A reduced model with variables chosen at least 33 out of the 100 times was used as a simpler alternative for comparison. This model, without cancer as a predictor, was also evaluated separately, considering that medical indications for THA surgery may look different for patients with cancer compared to cancer free patients. Univariable models with the ASA class, Charlson or Elixhauser co-morbidity indices were used for benchmarking, as well as a multivariable model with age and sex. Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots. Odds ratios for the final model were estimated with 95% confidence intervals.[@Ripley2002]

@@

The main and reduced models were no different regarding discriminative power, `r sprintf("%s versus %s", gsub(")", "", get_auc("Main model"), fixed = TRUE), gsub("(", "", get_auc("Reduced model"), fixed = TRUE))`. We therefore considered the reduced model as superior due to simplicity. Traditional models performed poorly with 95 % confidence intervals not above 0.7: ASA class `r get_auc("ASA")`, the Charlson co-morbitiy index `r get_auc("Charlson")`, and the Elixhauser com-morbidity index `r get_auc("Elixhauser")` (Figure \@ref(fig:rocs) left panel and \@ref(fig:aucci)).

@@

We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` better discriminated patients who died within 90 days after THA, from patients who survived, compared to traditional models with only ASA class or the Charlson or Elixhauser co-morbidity indices.

@@

A strength of the study is the nationwide design with data from national registers in both Sweden, England and Wales. The Swedish registers are valid with low proportions of missing data.[@Soderman2000; @shpr2018; @Soderman2001; @Ludvigsson2011] Some concerns have been raised regarding validity of the hospital episodes statistics (HES) database from England and Wales however.[@Spencer2012; @Williams2002; @Brennan2012] A systematic review found that the overall median diagnostic accuracy (comparing ICD-codes from HES to individual case notes) was 80%.[@Burns2012] Both in- and outpatient co-morbidity data was available for Sweden, but only in-patient data for England and Wales. We did not have acces to any data from primary care or nursing homes in any of the counties.

@@

>>>>

---
title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
author: Anne Garland`*`^1,2,3^, Erik Bulow`*`^2,4^, Erik Lenguerrand^5^, Ashley Blom^5^, Mark Wilkinson^6^, Adrian Sayers^5^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
date: '`r Sys.Date()`'
bibliography: H:/zotero_lib.bib
csl: the-lancet.csl
output: 
  bookdown::word_document2:
    reference_docx: article.dotx
    toc: false
    df_print: kable
---

@@

**OBJECTIVE:** Shared individualized decision making prior to total hip arthroplasty (THA) includes discussions of the risk of mortality. However, diagnosis code-based instruments such as the Charlson co-morbidity index are impossible to use in clinical practice, and the American Society for Anesthesiologists (ASA) class is imprecise. We thus developed a simple model to predict 90-day mortality after elective THA.

@@

**RESLUTS:** The unadjusted cumulative 90-day mortality was `r surv90d` in Sweden and `r njr_surv` in England and Wales. We propose a simple main effects model combining `r coefs_print_reduced`. This model had good discriminatory ability both internally `r get_auc("Reduced model")` and externally `r get_auc("NJR (Reduced model)")`, rendering it superioir to traditional models based on the Charlson `r get_auc("Charlson")`, and the Elixhauser co-morbidity indices `r get_auc("Elixhauser")`. Our model was well calibrated for predicted probabilities up to 5 %. A web calculator to aid clinical usage was published at `r shiny_url`.

@@

Pre-operative co-morbidity data have been used to develop clinically applicable prediction models for early mortality and adverse events such as surgical site infection in the context of THA surgery.[@Bozic2013a; @Bozic2013b; @Harris2018; @Harris2018a; @Inacio2016; @Price2019; @Harris2019] The first prediction model based on a European cohort was the Patient Decision support tool for joint remplacement in the UK (https://jointcalc.shef.ac.uk/). Due to limited, often single-center based samples, a lack of reported calibration abilities, the absence of external validation, and the use of extrapolation, there is still room for improved prediction models of early mortality.[@Manning2016]

@@

We used data from the Swedish Hip Arthroplasty Register (SHAR) for model derivation and internal validation of discrimination and calibration. The best model was then validated externally on patients from England and Wales recorded in the National Joint Registry for England, Wales, Northern Ireland, the Isle of Man and the States of Guernsey (NJR). We focused on patients with cemented THA due to osteoarthritis. Cementation is the most common fixation technique used in Sweden, thus used as inclusion criteria to increase heterogeneity. Only the last operated hip was accounted for in patients with bilateral THA, since those are shown to better resemble unilateral THAs.[@Bulow2020] (Figure \@ref(fig:flowchart)).

@@

The patient cohort was identified from the SHAR. The inclusion period started 2008 since this was the first year with the American Society for Anesthesiologists (ASA) class and Body Mass Index (BMI) systematically recorded. The last year was 2015, since we had access to co-morbidity from the national patient register up to this point. Deterministic data linkage was performed by 10-digit identity numbers, assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009] We used data from a variety of sources, as previously described by Cnudde et al.[@Cnudde2016] Age, sex, BMI, ASA class, type of hospital (university/county/rural/private) and year of surgery were collected from the SHAR, with a completeness of 96-98%.[@Karrholm2018] Data on educational level (low = up to 9 years/middle = 10-12 years/high = at least 12 years) was recorded for more than 98 % of the population with 85 % accuracy in the longitudinal integration database for health insurance and labor market studies (LISA) from Statistics Sweden.[@Ludvigsson2019] Civil status (married/un-married/divorced/widow[er]), were also collected from LISA. The Swedish National Patient Register (NPR) was used to assess co-morbidity during the year preceding index surgery. This register contains all relevant diagnoses coded by ICD-10, as well as dates of admission and discharge for in- and outpatient episodes in all private and public hospitals. Completeness for NPR is above 99 % and 85-95 % of all diagnostic codes are valid.[@Ludvigsson2011] Death dates were linked from the national population register.

@@

We used the Kaplan-Meier estimator to assess unadjusted mortality after cemented THA as background knowledge. Further analysis was based on logistic regression, since no censoring occurred in the Swedish cohort within the first 90 days, and we assumed the same was true for patients from England and Wales. We used a modelling procedure with bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020] Numeric variables (age and BMI) were normalized before modelling to have mean = 0 and standard deviation = 1. This way, the magnitude (absolute values) of their estimated coefficients would indicate variable importance on the same scale as the categorical variables. Co-morbidities recorded for at least one patient who died within 90 days, and one who did not, were included in the modelling process. 1,000 bootstrap samples were drawn from the observed data set.[@Austin2004] We used 10-fold cross validation for every bootstrap sample with a broad range of potential penalty values ($\lambda$:s) in a logistic LASSO model. We then only kept $\lambda$:s minimizing the mean cross-validated deviances in each sample. Those $\lambda$:s were used to estimate model coefficients for each potential predictor. The magnitude (absolute values) of those estimates were used as a measure of variable importance. Piece-wise linear regression was used to detect a break point where a significant drop in variable importance was observed. Potential predictors with variable importance above this break point were considered important and kept as model candidates. The whole process was repeated 100 times. Covariates that were selected at least once out of the 100 times were used in a main effects model of multivariable logistic regression without penalty, and without pre-normalization of numeric variables (main model). A reduced model with variables chosen at least 33 out of the 100 times was used as a simpler alternative for comparison. This model, without cancer as a predictor, was also evaluated separately, considering that medical indications for THA surgery may look different for patients with cancer compared to cancer free patients. Univariable models with the ASA class, Charlson or Elixhauser co-morbidity indices were used for benchmarking, as well as a multivariable model with age and sex. Each model including age was fitted three times, once with age as a main effect and twice with restricted cubic splines, either by two or three knots. Odds ratios for the final model were estimated with 95% confidence intervals.[@Ripley2002]

@@

The main and reduced models were no different regarding discriminative power, `r sprintf("%s versus %s", gsub(")", "", get_auc("Main model"), fixed = TRUE), gsub("(", "", get_auc("Reduced model"), fixed = TRUE))`. We therefore considered the reduced model as superior due to simplicity. Traditional models performed poorly with 95 % confidence intervals not above 0.7: the Charlson co-morbitiy index `r get_auc("Charlson")`, and the Elixhauser com-morbidity index `r get_auc("Elixhauser")` (Figure \@ref(fig:rocs) left panel and \@ref(fig:aucci)).

@@

We found that a multivariable main effects logistic regression model with `r coefs_print_reduced` better discriminated patients who died within 90 days after THA, from patients who survived, compared to traditional models with the Charlson or Elixhauser co-morbidity indices.

@@

A strength of the study is the nationwide design with data from national registers in both Sweden, England and Wales. The Swedish registers are valid with low proportions of missing data.[@Soderman2000; @Karrholm2018; @Soderman2001; @Ludvigsson2011] Some concerns have been raised regarding validity of the hospital episodes statistics (HES) database from England and Wales however.[@Spencer2012; @Williams2002; @Brennan2012] A systematic review found that the overall median diagnostic accuracy (comparing ICD-codes from HES to individual case notes) was 80%.[@Burns2012] Both in- and outpatient co-morbidity data was available for Sweden, but only in-patient data for England and Wales. We did not have acces to any data from primary care or nursing homes in any of the counties.

@@

We investigated wether the effect of age on mortality was non-linear using splines. A similar approach might be relevant for BMI as well, which we only modelled as a linear main effect. A U-shaped association between BMI categorized as underweight (below 18.5 kg/m^2^), normal (18.5 to 25 kg/m^2^), or obese (above 25 kg/m^2^) has been showed for example in England and Wales.[@Mouchti2018] Our recordings of BMI might be imprecise however.[@Stommel2009] Some h○ospitals might report actual measurements but it is unknown to what extent those data are self-estmiated by the patients. Therefore, we aimed to avoid overfitting BMI based on too elaborate modelling.
